Alternative Methods for Retaining Explicit and Finding Implicit Sharing in embedded DSLs

### 2022-12-16 paper/paper.tex
typos

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Lucas Dutton \and
Steven Gonder \and
Christopher Kumar Anand
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  Detection of sharing is known challenge for implementers of embedded domain
  specific languages (DSLs). There are many solutions, each with their
  advantages and drawbacks. Many solutions are based on observable sharing, that
  requires either a monadic interface or use of unsafe referencing, e.g.,
  Data.Reify. Monadic interfaces are consider unsuitable for domain experts, and
  the use of unsafe referencing leads to fragile software.

@@

Consider the naive DSL implemented as a Haskell data type:
\begin{minted}{haskell}
data Exp
  = Add Exp Exp
  | Variable String
  | Constant Int
\end{minted}
Expressions generate Abstract Syntax Trees (ASTs),
but consider this example,
\begin{minted}{haskell}
v0 = Variable "v0"
exp0 = Add v0 (Constant 0)
exp1 = Add exp0 exp0
\end{minted}
in which the expression  \mintinline{haskell}{exp0} is shared,
and will therefore be stored once in memory.
For large expressions with lots of sharing,
this can make a substantial difference.
However,
one of the first things the developer will do is write a pretty printer.
That recursive function will traverse the data structure as a tree,
and pretty print \mintinline{haskell}{exp0} twice.
This inefficiency is a problem for code generation,
and naive traversal of the AST does the opposite of common-subexpression elimination performed by a good optimizing compiler.
To avoid this,
rather than representing the code as an AST, 
we should use a DAG, retaining all of the sharing in the original DSL code.

@@

Nodes need to be uniquely identifiable, and shouldn't be a recursive data type,
such as
\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int
\end{minted}
The representation for the finally tagless
instance is then a wrapper around a State monad that holds the DAG being constructed in its state and
returns the current (top) \mintinline{haskell}{NodeID}:
\begin{minted}{haskell}
newtype DAG = DAG (BiMap Node) deriving Show

@@

When we wrap our State monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
\begin{minted}{haskell}
haskellSharing x y=
 let
   z = x + y
 in z + z
\end{minted}

@@

One solution to this issue is to integrate container types such as tuples and
lists into the DSL language. However doing this eliminates the
advantage of having an embedded language.
Manipulating tuple values will be
cumbersome, constantly requiring calls to custom implementations of \mintinline{haskell}{fst},\mintinline{haskell}{snd} etc. 
And for lists you'll lose access to built-in Haskell
list functionality.

@@

The heart of our problem is that whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
its unique identifier? We need some way to uniquely identify it outside the monad.

@@

To prevent confusion, we rename the hash-consing function to \mintinline{haskell}{triecons}, as follows
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to look up its corresponding byte string AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

@@

\subsection{Memory Limitations}
The byte string AST being built will itself suffer from lack of sharing. We're
essentially trading extra computation for extra memory. In our same
\mintinline{haskell}{addChains} example from Section \ref{limithashcons}, our
method now has exponential scaling in memory instead of computation. This can be
a good tradeoff, since memory is so plentiful in modern hardware, but still
presents an issue.

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from 
\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label. For safety purposes, we need to keep track of a table of
these labels and their corresponding ASTs, to make sure we don't use the same label for different ASTs.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagSubMap :: Map ByteString ByteString
               , dagMaxID :: Int
               } deriving Show

@@

subTInsert :: Maybe ByteString -> ByteString
           -> (Node, NodeID) -> State DAG ()
subTInsert Nothing  _ _  = return ()
subTInsert (Just s) sAST nodeID =
  do DAG trie subtMap _ <- get
     case Map.lookup sAST subtMap of
        Just sAST' -> if sAST == sAST'
                      then return ()
                      else error "tried to resubT"
        Nothing -> let cMap' = Map.insert sAST s subtMap
                       trie' = Trie.insert sAST nodeID trie
                  in modify (\dag -> dag { dagTrie = trie'
                                      , dagSubMap = cMap' })
\end{minted}
We need to make sure we don't attempt to insert the same substitution for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the State monad to prevent or modify the substitution.
In the toy example, compilation crashes, but we could catch an exception instead.
Either way it's up to the DSL
writer to ensure they don't reuse the same label.

@@

\section{Conclusion}
We have presented a method for constructing finally tagless style DSLs with
sharing detection, that allow for multiple-output DSL functions.
It also avoids the use of unsafe referencing as performed when doing observable sharing, c.f. \cite{gill:observablesharing}.

>>>>

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Retaining
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Lucas Dutton \and
Steven Gonder \and
Christopher Kumar Anand
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  Detection of sharing is a known challenge for implementers of embedded domain
  specific languages (DSLs). There are many solutions, each with their
  advantages and drawbacks. Many solutions are based on observable sharing, that
  requires either a monadic interface or use of unsafe referencing, e.g.,
  Data.Reify. Monadic interfaces are considered unsuitable for domain experts, and
  the use of unsafe referencing leads to fragile software.

@@

Consider the naive DSL implemented as a Haskell data type:
\begin{minted}{haskell}
data Exp
  = Add Exp Exp
  | Variable String
  | Constant Int
\end{minted}
Expressions generate Abstract Syntax Trees (ASTs),
but consider this example,
\begin{minted}{haskell}
v0 = Variable "v0"
exp0 = Add v0 (Constant 0)
exp1 = Add exp0 exp0
\end{minted}
in which the expression  \mintinline{haskell}{exp0} is shared,
and will therefore be stored once in memory.
For large expressions with lots of sharing,
this can make a substantial difference.

One of the first things the developer will do is write a pretty printer.
That recursive function will traverse the data structure as a tree,
and pretty print \mintinline{haskell}{exp0} twice.
This inefficiency is a real problem for code generation,
and naive traversal of the AST does the opposite of the common-subexpression elimination performed by a good optimizing compiler.
To avoid this,
rather than representing the code as an AST, 
we should use a DAG, retaining all of the sharing in the original DSL code.

@@

Nodes need to be uniquely identifiable, and shouldn't be a recursive data type,
such as
\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int
\end{minted}
The representation for the finally tagless
instance is then a wrapper around a state monad that holds the DAG being constructed in its state and
returns the current (top) \mintinline{haskell}{NodeID}:
\begin{minted}{haskell}
newtype DAG = DAG (BiMap Node) deriving Show

@@

When we wrap our state monad in finally tagless style, we lose some expected sharing. In the following code, the use of the let causes the computation $x + y$ to only occur once
\begin{minted}{haskell}
haskellSharing x y=
 let
   z = x + y
 in z + z
\end{minted}

@@

One solution to this issue is to integrate container types such as tuples and
lists into the DSL language. However doing this eliminates the
advantage of having an embedded language.
Manipulating tuple values will be
cumbersome, constantly requiring calls to custom implementations of \mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. 
And for lists you'll lose access to built-in Haskell
list functionality.

@@

The heart of our problem is that whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
its unique identifier. We need some way to uniquely identify it outside the monad.

@@

To prevent confusion, we name the hash-consing function in our method \mintinline{haskell}{triecons}:
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to find its corresponding serialized AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

@@

\subsection{Memory Limitations}
The byte string AST being built will itself suffer from lack of sharing. We're
essentially trading extra computation for extra memory. In our 
\mintinline{haskell}{addChains} example from Section~\ref{limithashcons}, our
method now has exponential scaling in memory instead of computation. This can be
a good tradeoff, since memory is so plentiful in modern hardware, but still
presents an issue.

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from Kiselyov 
\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label. For safety purposes, we need to keep track of a table of
these labels and their corresponding ASTs, to make sure we don't use the same label for different ASTs.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagSubMap :: Map ByteString ByteString
               , dagMaxID :: Int
               } deriving Show

@@

subTInsert :: Maybe ByteString -> ByteString
           -> (Node, NodeID) -> State DAG ()
subTInsert Nothing  _ _  = return ()
subTInsert (Just s) sAST nodeID =
  do DAG trie subtMap _ <- get
     case Map.lookup sAST subtMap of
        Just sAST' -> if sAST == sAST'
                      then return ()
                      else error "tried to resubT"
        Nothing -> let cMap' = Map.insert sAST s subtMap
                       trie' = Trie.insert sAST nodeID trie
                  in modify (\dag -> dag { dagTrie = trie'
                                      , dagSubMap = cMap' })
\end{minted}
We need to make sure we don't attempt to insert the same substitution for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the state monad to prevent or modify the substitution.
In the toy example, compilation crashes, but we could catch an exception instead.
Either way it's up to the DSL
writer to insure they don't reuse the same label.

@@

\section{Conclusion}
We have presented a method for constructing finally tagless style DSLs with
sharing detection, that allows for DSLs specifying hypergraphs (e.g., functions with multiple outputs).
It also avoids the use of unsafe referencing as performed when doing observable sharing, c.f. \cite{gill:observablesharing}.
